---
title: "Hierarchical Modeling With Animated Plots"
author: "Rob Kravec"
date: "5/3/2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
library(tidyverse)
library(stringr)
library(lme4)
library(knitr)
library(gganimate)
library(ggdist)
library(arm)
library(patchwork)
```

## Introduction

The famous statistician and founder of the sabermetrics movement in baseball,
Bill James, found that winning percentage 
($\frac{\text{wins}}{\text{games played}}$) has
a strong linear relationship with a simple function of a team's runs scored and
runs allowed in a given year. That relationship, coined as the "pythagorean 
expectation" was originally represented as 
$\frac{\text{runs scored}^{2}}{\text{runs scored}^{2} + \text{runs allowed}^{2}}$.
The exponents in this expression have since been amended to create an even 
stronger relationship with winning percentage (i.e., changing all powers of 2 
to ~ 1.83), and we will proceed with the amended version:
$\frac{\text{runs scored}^{1.83}}{\text{runs scored}^{1.83} + \text{runs allowed}^{1.83}}$

This pythagorean expectation can be used in retrospective analyses to see which
teams overperformed relative to their runs scored and runs allowed (i.e., 
pythagorean expectation < winning percentage) vs. which teams underperformed
(i.e, pythagorean expectation > winning percentage). One can then ask the 
question: Are there teams that chronically overperform or underperform? 

To assess this question, we'll use 20 Major League Baseball seasons' worth of
data from Fangraphs and build some simple models. Baseball (and sports in
general) provides ample opportunity to consider nested hierarchical structures 
(e.g., players within teams, teams within divisions, divisions within
leagues), so we'll evaluate whether these structures are helpful in our models. 

## EDA

First, we should be transparent about a few data processing steps:

- Fangraphs provides batting and pitching statistics separately. To create
a single dataframe with both runs scored and runs allowed, we simply download
and join two CSV files and get a 100% match rate on the join (by team and season)
- Four teams changed names during the 20 year period under consideration 
(2000-2019). To maintain a total of 30 unique teams for this analysis, previous
team names were re-coded to match current team names (e.g., Florida Marlins was
re-coded to Miami Marlins)
- The Houston Astros changed divisions in 2013, and our data reflects that
change (e.g., NL Central before 2013 and AL West starting in 2013)

```{r load-data}
# Load data files
batting <- read_csv('Batting2.csv')
pitching <- read_csv('Pitching2.csv')

# Keep only columns relevant to this analysis
batting_small <- batting %>% 
  mutate(Runs_scored = R) %>% 
  dplyr::select(Season, Team, Runs_scored)
pitching_small <- pitching %>% 
  mutate(Runs_allowed = R,
         Win_perc = W / (W + L)) %>% 
  dplyr::select(Season, Team, Runs_allowed, W, L, Win_perc)

# Join data
combined <- left_join(x = pitching_small, y = batting_small,
                      by = c("Team", "Season"))

# Recode some team names
combined$Team <- recode(combined$Team, FLA = "MIA")
combined$Team <- recode(combined$Team, ANA = "LAA")
combined$Team <- recode(combined$Team, MON = "WSN")
combined$Team <- recode(combined$Team, TBD = "TBR")

### Add relevant features
## Division
combined <- combined %>% 
  mutate(Division = case_when(
    (Team == "ATL" | Team == "PHI" | Team == "MIA" |
      Team == "WSN" | Team == "NYM") ~ "NL East",
    (Team == "CIN" | Team == "STL" | Team == "CHC" |
      Team == "MIL" | Team == "PIT") ~ "NL Central",
    (Team == "LAD" | Team == "SDP" | Team == "SFG" |
      Team == "COL" | Team == "ARI") ~ "NL West",
    (Team == "BOS" | Team == "BAL" | Team == "TOR" |
      Team == "TBR" | Team == "NYY") ~ "AL East",
    (Team == "MIN" | Team == "CHW" | Team == "CLE" |
      Team == "KCR" | Team == "DET") ~ "AL Central",
    (Team == "HOU" | Team == "LAA" | Team == "TEX" |
      Team == "SEA" | Team == "OAK") ~ "AL West"
  ),
  Team = as.factor(Team))
# Account for fact that Astros switched divisions in 2013
combined$Division <- ifelse(combined$Team == "HOU" & combined$Season < 2013,
                            "NL Central", combined$Division)

## League
combined$League <- str_sub(combined$Division, start = 1, end = 2)

## Pythagorean expectation
combined$Pythagorean <- 
  1 / (1 + (combined$Runs_allowed / combined$Runs_scored) ^ 1.83)

## Remove unnecessary tables
rm(batting, batting_small, pitching, pitching_small)
```

Jumping into our exploratory data analysis, we are first interested in 
confirming that the pythagorean expectation is as good as advertised in terms
of describing variance in winning percentage. Below, we see that this 
relationship is indeed very strong and linear. 

```{r aggregate-plot, fig.height = 2.5}
ggplot(data = combined, mapping = aes(x = Pythagorean, y = Win_perc)) +
  geom_point() + geom_smooth(method = 'lm', se = F) +
  labs(x = "Pythagorean expectation", y = "Winning percentage",
       title = "Winning Percentage vs. Pythagorean expectation") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

When looking at individual teams, however, we can see differences in 
the orientations of the best-fit lines. For instance, the New York Yankees and 
Toronto Blue Jays appear to have best-fit lines with smaller slopes than those
of the Kansas City Royals and Baltimore Orioles.

```{r team-plot}
ggplot(data = combined, mapping = aes(x = Pythagorean, y = Win_perc)) +
  geom_point() + geom_smooth(method = 'lm', se = F) +
  facet_wrap(~ Team) +
  labs(x = "Pythagorean expectation", y = "Winning percentage",
       title = "Winning Percentage vs. Pythagorean Expectation by Team") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

While we originally had aspirations of building a hierarchical model with
multiple layers of nesting structure, we see that the relationship between
winning percentage and pythagorean expectation is approximately identical 
across divisions. Though not pictured, the same can be said when aggregating the 
data even further to the league level (i.e., American League vs. National League).

```{r division-plot}
ggplot(data = combined, mapping = aes(x = Pythagorean, y = Win_perc)) +
  geom_point() + geom_smooth(method = 'lm', se = F) +
  facet_wrap(~ Division) +
  labs(x = "Pythagorean expectation", y = "Winning percentage",
       title = "Winning Percentage vs. pythagorean expectation by division") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r League-plot, eval = F}
ggplot(data = combined, mapping = aes(x = Pythagorean, y = Win_perc)) +
  geom_point() + geom_smooth(method = 'lm', se = F) +
  facet_wrap(~ League) +
  labs(x = "Pythagorean expectation", y = "Winning percentage",
       title = "Winning Percentage vs. Pythagorean expectation by league") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

## Models and results

As noted earlier in this analysis, our data has a natural grouping structure 
with multiple observations (Seasons) for each team, which belong to divisions,
which belong to leagues. Based on our EDA, we have reason to believe that
the relationship between winning percentage and pythagorean expectation may be
meaningfully different by team but likely not by division or league. For now,
let's consider only grouping by the `Team` variable.

As is common in hierarchical modeling contexts, we'll consider 3 different
model types and compare their results. To clarify notation used in model 
specifications, $y$ denotes winning percentage, $x$ denotes pythagorean expectation,
subscript $i$ corresponds to an observation, and subscript $j$ corresponds to 
a team.

- Pooled model: $y_i = \beta_0 + \beta_1x_i + \epsilon_{i}$
  - Lumps all data together and ignores the fact that any natural groupings
  exist
- Unpooled model: $y_{ij} = \beta_{0j} + \beta_{1j}x_{ij} + \epsilon_{ij}$
  - Fits a separate model for each team, ignoring any similarities
- Hierarchical model (shown below)
  - Allows for partial pooling of information across teams, pulling extreme
  values of coefficients toward the population average and reducing variance
  in coefficient estimates
  - In this case, we specify random intercepts and slopes by team

$$
\begin{aligned}
y_{ij} = \beta_0 + b_{0j} + \beta_1 + b_{1j} + \epsilon_{ij} \\
\begin{pmatrix} b_{0j} & b_{1j} \end{pmatrix} ^ T \sim N_2(0, \Sigma) \\
\epsilon_{ij} \sim N(0, \sigma^2) \\
b_{0j}, b_{1j} \perp \!\!\! \perp \epsilon_{ij}
\end{aligned}
$$

Lastly, before fitting any models, we note that some may object to us
considering winning percentage as a continuous numeric response, as opposed 
to modeling wins and losses with a generalized linear model, which would view
individual games as Bernoulli trials. We are comfortable proceeding with 
winning percentage as a continuous dependent variable because values are
entirely contained in the 0.25 - 0.75 range. As a result, we have little to fear
in terms of a linear regression predicting values of the response that are 
outside of the acceptable 0 - 1 range.

In an attempt to share results in a manner that is both clear and engaging, we
will make use of the `gganimate` package. The slope estimate for the pooled
model is shown with a vertical dashed red line, while 95% confidence intervals
for the team-level slope estimates are shown for the unpooled and hierarchical
models. The subtitle of the plot denotes which coefficient estimates (unpooled
or hierarchical) are being shown at any one time.

```{r pooled-model}
mod1 <- lm(Win_perc ~ Pythagorean, data = combined)
pooled_df <- data.frame(Intercept = coef(mod1)[[1]],
                      Slope = coef(mod1)[[2]],
                      Slope_se = summary(mod1)[[4]][4]) %>% 
  mutate(Slope_lower = Slope - 2 * Slope_se,
         Slope_upper = Slope + 2 * Slope_se)
```

```{r unpooled-models}
# Write function for unpooled regression (by team)
unpooled <- function(team) {
  dat <- combined %>% 
    filter(Team == team)
  mod <- lm(Win_perc ~ Pythagorean, data = dat)
  df <- data.frame(Team = team,
                   Intercept = coef(mod)[[1]],
                   Slope = coef(mod)[[2]],
                   Slope_se = summary(mod)[[4]][4]) %>% 
    mutate(Slope_lower = Slope - 2 * Slope_se,
           Slope_upper = Slope + 2 * Slope_se)
  return(df)
}

# Define unique teams
teams <- unique(combined$Team)

# Find unpooled regression coefficients
unpooled_df <- map_df(.x = teams, .f = unpooled)

# Add identifier that will be helpful for animation
unpooled_df <- unpooled_df %>% 
  mutate(Type = "Unpooled")
```

```{r plot-pooled-vs-unpooled}
unpooled_plot <- ggplot() +
  geom_vline(mapping = aes(xintercept = pooled_df$Slope, 
                           color = "Pooled estimate"),
                           linetype = "dashed") +
  geom_pointinterval(data = unpooled_df, 
                     mapping = aes(x = Slope, xmin = Slope_lower,
                                   xmax = Slope_upper,
                                   y = fct_reorder(Team, Slope))) +
  labs(x = "Slope estimate", title = "Unpooled slope estimates",
       color = "Legend") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none") +
  coord_cartesian(xlim = c(0.2, 1.7))
```

```{r team-random-effects}
# Fit model with random slopes and intercepts by Team
mod2 <- lmer(Win_perc ~ (1 + Pythagorean | Team), data = combined)

# Extract random effects and slope standard error
std_errs <- se.ranef(mod2)[[1]][,2]
slope_res <- ranef(mod2)[[1]][,2]
intercept_res <- ranef(mod2)[[1]][,1]
teams <- unique(combined$Team) %>% sort()

# Create data frame that will be helpful for plotting
mod2_df <- data.frame(Team = teams,
                      Intercept = intercept_res,
                      Slope = slope_res,
                      Slope_se = std_errs,
                      Slope_lower = slope_res - 2 * std_errs,
                      Slope_upper = slope_res + 2 * std_errs,
                      Type = "Hierarchical")

# Plot result
hierarchical_plot <- ggplot(data = mod2_df) +
  geom_vline(mapping = aes(xintercept = pooled_df$Slope, 
                           color = "Pooled estimate"),
                           linetype = "dashed") +
  geom_pointinterval(mapping = aes(x = Slope, xmin = Slope_lower,
                                   xmax = Slope_upper,
                                   y = fct_reorder(Team, Slope))) +
  labs(title = "Multilevel model slope estimates",
       color = "Legend") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  coord_cartesian(xlim = c(0.2, 1.7))

# Display patchwork plot
#unpooled_plot + hierarchical_plot
```

```{r animated-plot}
# Create data frames for animated plotting
anim_df <- rbind(unpooled_df, mod2_df)
pooled_df <- rbind(pooled_df, pooled_df) %>% 
  mutate(Type = c("Unpooled", "Hierarchical"))

# Create animated plot
anim_plot <- ggplot() +
  geom_vline(mapping = aes(xintercept = pooled_df$Slope, 
                           color = "Pooled estimate"),
                           linetype = "dashed") +
  geom_pointinterval(data = anim_df, 
                     mapping = aes(x = Slope, xmin = Slope_lower,
                                   xmax = Slope_upper,
                                   y = fct_reorder(Team, Slope))) +
  transition_states(Type, transition_length = 3, state_length = 1) +
  labs(x = "Slope estimate", title = "Slope estimates by model type",
       subtitle = "Model type: {closest_state}",
       color = "Legend") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.y = element_blank(),
        axis.title.x = element_blank())

# Display animation
animate(anim_plot, fps = 10, end_pause = 25)
```

```{r check-for-larger-ci}
unpooled_df2 <- unpooled_df %>% 
  mutate(ci_width_unpooled = Slope_upper - Slope_lower) %>% 
  dplyr::select(Team, Slope, ci_width_unpooled)
mod2_df2 <- mod2_df %>% 
  mutate(ci_width_hierarchy = Slope_upper - Slope_lower) %>% 
  dplyr::select(Team, ci_width_hierarchy)
compare_ci <- left_join(x = unpooled_df2, y = mod2_df2, by = "Team") %>% 
  mutate(ind = ifelse(ci_width_hierarchy > ci_width_unpooled, 1, 0))
```

On aggregate, we observe the expected result. The confidence intervals of the
slope estimates tend to shrink in the hierarchical model (relative to the
unpooled models), and the point estimates tend to move toward the population
estimate. As a result, a slightly greater number of the team-level slope 
confidence intervals contains the population estimate under the hierarchical
model. These phenomena are governed by the partial pooling of information
that we discussed earlier.

However, we also observe a couple of unexpected (and slightly troubling) results:

- For a non-negligible number of teams (10 of 30), the confidence interval
actually increases slightly in the hierarchical model
- Slope point estimates for some teams actually move further away from the 
population estimate and/or cross from one side of the population estimate over
to the other

It is my belief that these unintuitive results stem from the hierarchical 
model's attempt to balance slope estimates (and their corresponding variances)
across teams. Typically, this balancing would result in a shrinkage in variance
across the board and gradual movement toward the observed population mean. 
However, in this case, the extremely linear relationship between winning 
percentage and pythagorean expectation may be causing havoc with this 
re-balancing process. An additional proof point for this hypothesis can be
gleaned from a Normal Q-Q plot of the hierarchical model residuals, which
shows that although the residuals appear to have a "normal" shape, the tails
of the distribution are WAY shorter than expected. In essence, winning 
percentage is being predicted too well by the pythagorean expectation to 
satisfy the assumptions in the normal model.

```{r qq-plot, fig.height = 3.5}
qqnorm(resid(mod2))
```

Often, these types of model issues are exacerbated when sample sizes are low. 
As an educational experiment, we reproduce the same animated plot of slope
estimates below with 10 seasons of data per team instead of the original 20.
As expected, we observe similar results compared to the data with 20 seasons.
On aggregate, we see a reduction in confidence interval width and shrinkage
of estimates toward the population estimate. However, the "unintuitive" results
are more perceptible. To highlight two example, the confidence intervals for the 
New York Mets and Los Angeles Angels noticeably increase as the animation moves
from the unpooled estimates to the hierarchical model estimates.

```{r filter-10-seasons}
combined_2 <- combined %>% 
  filter(Season >= 2010)
```

```{r 10-seasons-plot}
# Pooled model
mod1_2 <- lm(Win_perc ~ Pythagorean, data = combined_2)
pooled_df_2 <- data.frame(Intercept = coef(mod1_2)[[1]],
                      Slope = coef(mod1_2)[[2]],
                      Slope_se = summary(mod1_2)[[4]][4]) %>% 
  mutate(Slope_lower = Slope - 2 * Slope_se,
         Slope_upper = Slope + 2 * Slope_se)

# Unpooled model
unpooled_2 <- function(team) {
  dat <- combined_2 %>% 
    filter(Team == team)
  mod <- lm(Win_perc ~ Pythagorean, data = dat)
  df <- data.frame(Team = team,
                   Intercept = coef(mod)[[1]],
                   Slope = coef(mod)[[2]],
                   Slope_se = summary(mod)[[4]][4]) %>% 
    mutate(Slope_lower = Slope - 2 * Slope_se,
           Slope_upper = Slope + 2 * Slope_se)
  return(df)
}

# Find unpooled regression coefficients
unpooled_df_2 <- map_df(.x = teams, .f = unpooled_2) %>% 
  mutate(Type = "Unpooled")

# Hierarchical model
mod2_2 <- lmer(Win_perc ~ (1 + Pythagorean | Team), data = combined_2)
std_errs_2 <- se.ranef(mod2_2)[[1]][,2]
slope_res_2 <- ranef(mod2_2)[[1]][,2]
intercept_res_2 <- ranef(mod2_2)[[1]][,1]
teams_2 <- unique(combined_2$Team) %>% sort()
mod2_df_2 <- data.frame(Team = teams_2,
                      Intercept = intercept_res_2,
                      Slope = slope_res_2,
                      Slope_se = std_errs_2,
                      Slope_lower = slope_res_2 - 2 * std_errs_2,
                      Slope_upper = slope_res_2 + 2 * std_errs_2,
                      Type = "Hierarchical")

# Animated plot
anim_df_2 <- rbind(unpooled_df_2, mod2_df_2)
pooled_df_2 <- rbind(pooled_df_2, pooled_df_2) %>% 
  mutate(Type = c("Unpooled", "Hierarchical"))

# Create animated plot
anim_plot_2 <- ggplot() +
  geom_vline(mapping = aes(xintercept = pooled_df_2$Slope, 
                           color = "Pooled estimate"),
                           linetype = "dashed") +
  geom_pointinterval(data = anim_df_2, 
                     mapping = aes(x = Slope, xmin = Slope_lower,
                                   xmax = Slope_upper,
                                   y = fct_reorder(Team, Slope))) +
  transition_states(Type, transition_length = 3, state_length = 1) +
  labs(x = "Slope estimate", title = "Slope estimates by model type",
       subtitle = "Model type: {closest_state}",
       color = "Legend") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.y = element_blank(),
        axis.title.x = element_blank())

# Display animation
animate(anim_plot_2, fps = 10, end_pause = 25)
```

## Discussion

Initially, we were drawn to this dataset for two main reasons: (1) nested
hierarchical structure and (2) simplicity that would allow for demonstration of
key hierarchical modeling concepts. On both accounts, we encountered some 
surprises. 

It turns out that the `Division` and `League` add almost zero additional 
information to a hierarchical model of winning percentage vs. pythagorean
expectation (first seen in EDA and confirmed by fitting additional models that
were not discussed). As a result, `Team` was the only grouping variable that
we analyzed in any detail.

Additionally, the very strong linear relationship between winning percentage
and pythagorean expectation complicated what was expected to be a simple,
textbook illustration of partial pooling in a hierarchical model. Based on
this complication, we decided to evaluate the difference in modeling outcomes
for a dataset with 20 seasons vs. 10 seasons and found that erratic hierarchical
model behavior was more evident for the smaller sample size. This outcome 
serves as a good reminder that skepticism about model results and a commitment 
to performing model diagnostics are essential tools for data analysis.

## References

- Wikipedia [article](https://en.wikipedia.org/wiki/Pythagorean_expectation)
on the pythagorean expectation, a key variable in my analysis
- Fangraphs [website](https://www.fangraphs.com/leaders.aspx?pos=all&stats=pit&lg=all&qual=0&type=0&season=2019&month=0&season1=2000&ind=1&team=0,ts&rost=0&age=0&filter=&players=0&startdate=&enddate=):
Data source for the analysis conducted in this "tool." The search terms listed 
are for one of the two files used to create the final dataset. The other file
can be generated by selecting "Batting" instead of "Pitching" among the options
on the website
- Course [website](http://www2.stat.duke.edu/courses/Fall20/sta523/) for Duke's
STA 523 course in Fall 2020: I referred back to course notes on `gganimate` 
to produce the animated plots of slope confidence intervals by team for the
unpooled and hierarchical models

## Code appendix

```{r setup2, eval = F, echo = T}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
library(tidyverse)
library(stringr)
library(lme4)
library(knitr)
library(gganimate)
library(ggdist)
library(arm)
library(patchwork)
```

```{r load-data2, eval = F, echo = T}
# Load data files
batting <- read_csv('Batting2.csv')
pitching <- read_csv('Pitching2.csv')

# Keep only columns relevant to this analysis
batting_small <- batting %>% 
  mutate(Runs_scored = R) %>% 
  dplyr::select(Season, Team, Runs_scored)
pitching_small <- pitching %>% 
  mutate(Runs_allowed = R,
         Win_perc = W / (W + L)) %>% 
  dplyr::select(Season, Team, Runs_allowed, W, L, Win_perc)

# Join data
combined <- left_join(x = pitching_small, y = batting_small,
                      by = c("Team", "Season"))

# Recode some team names
combined$Team <- recode(combined$Team, FLA = "MIA")
combined$Team <- recode(combined$Team, ANA = "LAA")
combined$Team <- recode(combined$Team, MON = "WSN")
combined$Team <- recode(combined$Team, TBD = "TBR")

### Add relevant features
## Division
combined <- combined %>% 
  mutate(Division = case_when(
    (Team == "ATL" | Team == "PHI" | Team == "MIA" |
      Team == "WSN" | Team == "NYM") ~ "NL East",
    (Team == "CIN" | Team == "STL" | Team == "CHC" |
      Team == "MIL" | Team == "PIT") ~ "NL Central",
    (Team == "LAD" | Team == "SDP" | Team == "SFG" |
      Team == "COL" | Team == "ARI") ~ "NL West",
    (Team == "BOS" | Team == "BAL" | Team == "TOR" |
      Team == "TBR" | Team == "NYY") ~ "AL East",
    (Team == "MIN" | Team == "CHW" | Team == "CLE" |
      Team == "KCR" | Team == "DET") ~ "AL Central",
    (Team == "HOU" | Team == "LAA" | Team == "TEX" |
      Team == "SEA" | Team == "OAK") ~ "AL West"
  ),
  Team = as.factor(Team))
# Account for fact that Astros switched divisions in 2013
combined$Division <- ifelse(combined$Team == "HOU" & combined$Season < 2013,
                            "NL Central", combined$Division)

## League
combined$League <- str_sub(combined$Division, start = 1, end = 2)

## Pythagorean expectation
combined$Pythagorean <- 
  1 / (1 + (combined$Runs_allowed / combined$Runs_scored) ^ 1.83)

## Remove unnecessary tables
rm(batting, batting_small, pitching, pitching_small)
```

```{r aggregate-plot2, eval = F, echo = T}
ggplot(data = combined, mapping = aes(x = Pythagorean, y = Win_perc)) +
  geom_point() + geom_smooth(method = 'lm', se = F) +
  labs(x = "Pythagorean expectation", y = "Winning percentage",
       title = "Winning Percentage vs. Pythagorean expectation") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r team-plot2, eval = F, echo = T}
ggplot(data = combined, mapping = aes(x = Pythagorean, y = Win_perc)) +
  geom_point() + geom_smooth(method = 'lm', se = F) +
  facet_wrap(~ Team) +
  labs(x = "Pythagorean expectation", y = "Winning percentage",
       title = "Winning Percentage vs. Pythagorean Expectation by Team") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r division-plot2, eval = F, echo = T}
ggplot(data = combined, mapping = aes(x = Pythagorean, y = Win_perc)) +
  geom_point() + geom_smooth(method = 'lm', se = F) +
  facet_wrap(~ Division) +
  labs(x = "Pythagorean expectation", y = "Winning percentage",
       title = "Winning Percentage vs. pythagorean expectation by division") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r League-plot2, eval = F, echo = T}
ggplot(data = combined, mapping = aes(x = Pythagorean, y = Win_perc)) +
  geom_point() + geom_smooth(method = 'lm', se = F) +
  facet_wrap(~ League) +
  labs(x = "Pythagorean expectation", y = "Winning percentage",
       title = "Winning Percentage vs. Pythagorean expectation by league") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r pooled-model2, eval = F, echo = T}
mod1 <- lm(Win_perc ~ Pythagorean, data = combined)
pooled_df <- data.frame(Intercept = coef(mod1)[[1]],
                      Slope = coef(mod1)[[2]],
                      Slope_se = summary(mod1)[[4]][4]) %>% 
  mutate(Slope_lower = Slope - 2 * Slope_se,
         Slope_upper = Slope + 2 * Slope_se)
```

```{r unpooled-models2, eval = F, echo = T}
# Write function for unpooled regression (by team)
unpooled <- function(team) {
  dat <- combined %>% 
    filter(Team == team)
  mod <- lm(Win_perc ~ Pythagorean, data = dat)
  df <- data.frame(Team = team,
                   Intercept = coef(mod)[[1]],
                   Slope = coef(mod)[[2]],
                   Slope_se = summary(mod)[[4]][4]) %>% 
    mutate(Slope_lower = Slope - 2 * Slope_se,
           Slope_upper = Slope + 2 * Slope_se)
  return(df)
}

# Define unique teams
teams <- unique(combined$Team)

# Find unpooled regression coefficients
unpooled_df <- map_df(.x = teams, .f = unpooled)

# Add identifier that will be helpful for animation
unpooled_df <- unpooled_df %>% 
  mutate(Type = "Unpooled")
```

```{r plot-pooled-vs-unpooled2, eval = F, echo = T}
unpooled_plot <- ggplot() +
  geom_vline(mapping = aes(xintercept = pooled_df$Slope, 
                           color = "Pooled estimate"),
                           linetype = "dashed") +
  geom_pointinterval(data = unpooled_df, 
                     mapping = aes(x = Slope, xmin = Slope_lower,
                                   xmax = Slope_upper,
                                   y = fct_reorder(Team, Slope))) +
  labs(x = "Slope estimate", title = "Unpooled slope estimates",
       color = "Legend") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none") +
  coord_cartesian(xlim = c(0.2, 1.7))
```

```{r team-random-effects2, eval = F, echo = T}
# Fit model with random slopes and intercepts by Team
mod2 <- lmer(Win_perc ~ (1 + Pythagorean | Team), data = combined)

# Extract random effects and slope standard error
std_errs <- se.ranef(mod2)[[1]][,2]
slope_res <- ranef(mod2)[[1]][,2]
intercept_res <- ranef(mod2)[[1]][,1]
teams <- unique(combined$Team) %>% sort()

# Create data frame that will be helpful for plotting
mod2_df <- data.frame(Team = teams,
                      Intercept = intercept_res,
                      Slope = slope_res,
                      Slope_se = std_errs,
                      Slope_lower = slope_res - 2 * std_errs,
                      Slope_upper = slope_res + 2 * std_errs,
                      Type = "Hierarchical")

# Plot result
hierarchical_plot <- ggplot(data = mod2_df) +
  geom_vline(mapping = aes(xintercept = pooled_df$Slope, 
                           color = "Pooled estimate"),
                           linetype = "dashed") +
  geom_pointinterval(mapping = aes(x = Slope, xmin = Slope_lower,
                                   xmax = Slope_upper,
                                   y = fct_reorder(Team, Slope))) +
  labs(title = "Multilevel model slope estimates",
       color = "Legend") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  coord_cartesian(xlim = c(0.2, 1.7))
```

```{r animated-plot2, eval = F, echo = T}
# Create data frames for animated plotting
anim_df <- rbind(unpooled_df, mod2_df)
pooled_df <- rbind(pooled_df, pooled_df) %>% 
  mutate(Type = c("Unpooled", "Hierarchical"))

# Create animated plot
anim_plot <- ggplot() +
  geom_vline(mapping = aes(xintercept = pooled_df$Slope, 
                           color = "Pooled estimate"),
                           linetype = "dashed") +
  geom_pointinterval(data = anim_df, 
                     mapping = aes(x = Slope, xmin = Slope_lower,
                                   xmax = Slope_upper,
                                   y = fct_reorder(Team, Slope))) +
  transition_states(Type, transition_length = 3, state_length = 1) +
  labs(x = "Slope estimate", title = "Slope estimates by model type",
       subtitle = "Model type: {closest_state}",
       color = "Legend") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.y = element_blank(),
        axis.title.x = element_blank())

# Display animation
animate(anim_plot, fps = 10, end_pause = 25)
```

```{r check-for-larger-ci2, eval = F, echo = T}
unpooled_df2 <- unpooled_df %>% 
  mutate(ci_width_unpooled = Slope_upper - Slope_lower) %>% 
  dplyr::select(Team, Slope, ci_width_unpooled)
mod2_df2 <- mod2_df %>% 
  mutate(ci_width_hierarchy = Slope_upper - Slope_lower) %>% 
  dplyr::select(Team, ci_width_hierarchy)
compare_ci <- left_join(x = unpooled_df2, y = mod2_df2, by = "Team") %>% 
  mutate(ind = ifelse(ci_width_hierarchy > ci_width_unpooled, 1, 0))
```

```{r qq-plot2, eval = F, echo = T}
qqnorm(resid(mod2))
```

```{r filter-10-seasons2, eval = F, echo = T}
combined_2 <- combined %>% 
  filter(Season >= 2010)
```

```{r 10-seasons-plot2, eval = F, echo = T}
# Pooled model
mod1_2 <- lm(Win_perc ~ Pythagorean, data = combined_2)
pooled_df_2 <- data.frame(Intercept = coef(mod1_2)[[1]],
                      Slope = coef(mod1_2)[[2]],
                      Slope_se = summary(mod1_2)[[4]][4]) %>% 
  mutate(Slope_lower = Slope - 2 * Slope_se,
         Slope_upper = Slope + 2 * Slope_se)

# Unpooled model
unpooled_2 <- function(team) {
  dat <- combined_2 %>% 
    filter(Team == team)
  mod <- lm(Win_perc ~ Pythagorean, data = dat)
  df <- data.frame(Team = team,
                   Intercept = coef(mod)[[1]],
                   Slope = coef(mod)[[2]],
                   Slope_se = summary(mod)[[4]][4]) %>% 
    mutate(Slope_lower = Slope - 2 * Slope_se,
           Slope_upper = Slope + 2 * Slope_se)
  return(df)
}

# Find unpooled regression coefficients
unpooled_df_2 <- map_df(.x = teams, .f = unpooled_2) %>% 
  mutate(Type = "Unpooled")

# Hierarchical model
mod2_2 <- lmer(Win_perc ~ (1 + Pythagorean | Team), data = combined_2)
std_errs_2 <- se.ranef(mod2_2)[[1]][,2]
slope_res_2 <- ranef(mod2_2)[[1]][,2]
intercept_res_2 <- ranef(mod2_2)[[1]][,1]
teams_2 <- unique(combined_2$Team) %>% sort()
mod2_df_2 <- data.frame(Team = teams_2,
                      Intercept = intercept_res_2,
                      Slope = slope_res_2,
                      Slope_se = std_errs_2,
                      Slope_lower = slope_res_2 - 2 * std_errs_2,
                      Slope_upper = slope_res_2 + 2 * std_errs_2,
                      Type = "Hierarchical")

# Animated plot
anim_df_2 <- rbind(unpooled_df_2, mod2_df_2)
pooled_df_2 <- rbind(pooled_df_2, pooled_df_2) %>% 
  mutate(Type = c("Unpooled", "Hierarchical"))

# Create animated plot
anim_plot_2 <- ggplot() +
  geom_vline(mapping = aes(xintercept = pooled_df_2$Slope, 
                           color = "Pooled estimate"),
                           linetype = "dashed") +
  geom_pointinterval(data = anim_df_2, 
                     mapping = aes(x = Slope, xmin = Slope_lower,
                                   xmax = Slope_upper,
                                   y = fct_reorder(Team, Slope))) +
  transition_states(Type, transition_length = 3, state_length = 1) +
  labs(x = "Slope estimate", title = "Slope estimates by model type",
       subtitle = "Model type: {closest_state}",
       color = "Legend") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.y = element_blank(),
        axis.title.x = element_blank())

# Display animation
animate(anim_plot_2, fps = 10, end_pause = 25)
```
